My title

### 2024-09-23 paper/paper.qmd
paper charts added

# Display the simplified geometry data
head(data_geometry) %>%
  kable("latex", booktabs = TRUE, col.names = c("Simplified Geometry")) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  column_spec(1, width = "40em")  # Adjust the width as needed
```

@@

# Display the first few rows of the cleaned coordinate data
head(coordinates_data) %>%
  kable("latex", booktabs = TRUE, col.names = c("Longitude", "Latitude", "ID"))
```

@@

# Step 5: Display the summarized data in a table
tail(upgraded_summary) %>%
  kable("latex", booktabs = TRUE, col.names = c("Year", "Number of Upgraded Lanes"))
```

@@

# Step 5: Display only the first 6 rows of the summarized data in a table
head(installed_summary) %>%
  kable("latex", booktabs = TRUE, col.names = c("Year", "Number of Installed Bikeways"))
```

@@

# Step 5: Display the first 6 rows of the cleaned and classified data
head(final_lane_data) %>%
  kable("latex", booktabs = TRUE, col.names = c("Installed Year", "Upgraded Year", "Comfort Level"))
```

@@

```{r}
#| label: fig-one
#| fig-cap: one
#| echo: false

@@

```{r}
#| label: fig-two
#| fig-cap: two
#| echo: false
#| warning: false
#| message: false

@@

@@

Our results are summarized in @tbl-modelresults.

>>>>

# Display the simplified geometry data
head(data_geometry) %>%
  kable("latex", booktabs = TRUE, col.names = c("Simplified Geometry")) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  column_spec(1, width = "40em")  # Adjust the width as needed

@@

# Display the first few rows of the cleaned coordinate data
head(coordinates_data) %>%
  kable("latex", booktabs = TRUE, col.names = c("Longitude", "Latitude", "ID"))

@@

# Step 5: Display the summarized data in a table
tail(upgraded_summary) %>%
  kable("latex", booktabs = TRUE, col.names = c("Year", "Number of Upgraded Lanes"))

@@

# Step 5: Display only the first 6 rows of the summarized data in a table
head(installed_summary) %>%
  kable("latex", booktabs = TRUE, col.names = c("Year", "Number of Installed Bikeways"))

@@

# Step 5: Display the first 6 rows of the cleaned and classified data
head(final_lane_data) %>%
  kable("latex", booktabs = TRUE, col.names = c("Installed Year", "Upgraded Year", "Comfort Level"))

@@

```{r}
#| label: fig-installed
#| fig-cap: "Toronto Bikeways Installed by Year"
#| fig-subcap: ["Number", "Trend"]
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2
#| layout-nrow: 1

# Load the data
installed_data <- read_csv(here("data", "analysis_data", "installed_data.csv"))

# Generate the bar chart for "Number of Bikeways Installed by Year"
ggplot(installed_data, aes(x = INSTALLED, y = num_bikeways)) +
  geom_bar(stat = "identity", fill = "darkgrey") +  # Bar chart
  labs(x = "Year of Installation",
       y = "Number of Bikeways") +
  scale_x_continuous(breaks = seq(min(installed_data$INSTALLED), max(installed_data$INSTALLED), by = 1)) +  # Display all years
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotates x-axis labels for readability

# Generate the line chart with trend line for "Number of Bikeways Installed by Year"
ggplot(installed_data, aes(x = INSTALLED, y = num_bikeways)) +
  geom_line(aes(group = 1), color = "black", size = 1) +  # Adds a line connecting each point
  geom_point(color = "black", size = 2) +  # Adds points at each data value
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Adds a linear trend line
  labs(x = "Year of Installation",
       y = "Number of Bikeways") +
  scale_x_continuous(breaks = seq(min(installed_data$INSTALLED), max(installed_data$INSTALLED), by = 1)) +  # Display all years
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotates x-axis labels for readability

@@

```{r}
#| label: fig-upgraded
#| fig-cap: "Toronto Bikeways Upgraded by Year"
#| fig-subcap: ["Number", "Trend"]
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2
#| layout-nrow: 1

# Load the data
upgraded_data <- read_csv(here("data", "analysis_data", "upgraded_data.csv"))

# Generate the bar chart for "Number of Bikeways Upgraded by Year"
ggplot(upgraded_data, aes(x = UPGRADED, y = num_lanes)) +
  geom_bar(stat = "identity", fill = "darkgrey") +  # Bar chart
  labs(x = "Year of Upgrade",
       y = "Number of Bikeways") +
  scale_x_continuous(breaks = seq(min(upgraded_data$UPGRADED), max(upgraded_data$UPGRADED), by = 1)) +  # Display all years
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotates x-axis labels for readability

# Generate the line chart with trend line for "Number of Bikeways Upgraded by Year"
ggplot(upgraded_data, aes(x = UPGRADED, y = num_lanes)) +
  geom_line(aes(group = 1), color = "black", size = 1) +  # Adds a line connecting each point
  geom_point(color = "black", size = 2) +  # Adds points at each data value
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Adds a linear trend line
  labs(x = "Year of Upgrade",
       y = "Number of Bikeways") +
  scale_x_continuous(breaks = seq(min(upgraded_data$UPGRADED), max(upgraded_data$UPGRADED), by = 1)) +  # Display all years
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotates x-axis labels for readability

@@

```{r}
#| label: fig-comfort
#| fig-cap: "Toronto Bikeway Comfort Level Distribution"
#| fig-subcap: ["All levels", "Low Comfort", "Moderate Comfort", "High Comfort"]
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2
#| layout-nrow: 2

# Load the data
lane_type_data <- read_csv(here("data", "analysis_data", "lane_type_data.csv"))

# Combine the 'INSTALLED' and 'UPGRADED' years into one column for better visualization
lane_type_years <- lane_type_data %>%
  pivot_longer(cols = c(INSTALLED, UPGRADED), names_to = "Year_Type", values_to = "Year") %>%
  filter(!is.na(Year)) %>%  # Remove NA values in the 'Year' column
  group_by(Comfort_Level) %>%
  summarise(count = n())  # Count the occurrences of each Comfort Level

# Calculate percentages
lane_type_years <- lane_type_years %>%
  mutate(percentage = round((count / sum(count)) * 100, 2))  # Add percentage column

# Define a more contrasting Morandi-inspired color scheme
contrasting_morandi_colors <- c(
  "High Comfort" = "#B39B8E",   # Muted brownish-pink
  "Moderate Comfort" = "#8A7967", # Stronger grey-brown
  "Low Comfort" = "#6E7062"     # Dark muted gray-green
)

# Generate the pie chart with border lines and larger text/legend
ggplot(lane_type_years, aes(x = "", y = percentage, fill = Comfort_Level)) +
  geom_bar(stat = "identity", width = 1, color = "black") +  # Add black border lines around pie sections
  coord_polar("y", start = 0) +
  geom_text(aes(label = paste0(percentage, "%")), 
            position = position_stack(vjust = 0.5), 
            size = 6,   # Increase text size for percentage labels
            fontface = "bold") +  # Make percentage text bold
  scale_fill_manual(values = contrasting_morandi_colors) +  # Apply the more contrasting Morandi colors
  labs(fill = "Comfort Level") +
  theme_void() +  # Remove axis, grid lines, and background
  theme(legend.position = "right", 
        legend.title = element_text(size = 16),  # Increase legend title size
        legend.text = element_text(size = 14))  # Increase legend text size

# Filter data for 'Low Comfort' lanes and categorize as 'Upgraded' or 'Not Upgraded'
low_comfort_data <- lane_type_data %>%
  filter(Comfort_Level == "Low Comfort") %>%
  mutate(
    Upgrade_Status = ifelse(is.na(UPGRADED), "Not Upgraded", "Upgraded")
  ) %>%
  group_by(Upgrade_Status) %>%
  summarise(count = n())  # Count the occurrences of each category

# Calculate percentages
low_comfort_data <- low_comfort_data %>%
  mutate(percentage = round((count / sum(count)) * 100, 2))  # Add percentage column

# Define colors for 'Upgraded' and 'Not Upgraded' categories
upgrade_colors_low <- c(
  "Upgraded" = "#8A7967",  # Gray-brown for upgraded
  "Not Upgraded" = "#B39B8E"  # Muted pinkish tone for not upgraded
)

# Generate the pie chart with borders and larger text/legend for Low Comfort lanes
ggplot(low_comfort_data, aes(x = "", y = percentage, fill = Upgrade_Status)) +
  geom_bar(stat = "identity", width = 1, color = "black") +  # Add black border lines around pie sections
  coord_polar("y", start = 0) +
  geom_text(aes(label = paste0(percentage, "%")), 
            position = position_stack(vjust = 0.5), 
            size = 6,   # Increase text size for percentage labels
            fontface = "bold") +  # Make percentage text bold
  scale_fill_manual(values = upgrade_colors_low) +  # Apply colors for Upgraded vs Not Upgraded
  labs(fill = "Upgrade Status") +
  theme_void() +  # Remove axis, grid lines, and background
  theme(legend.position = "right", 
        legend.title = element_text(size = 16),  # Increase legend title size
        legend.text = element_text(size = 14))  # Increase legend text size

# Filter data for 'Moderate Comfort' lanes and categorize as 'Upgraded' or 'Not Upgraded'
moderate_comfort_data <- lane_type_data %>%
  filter(Comfort_Level == "Moderate Comfort") %>%
  mutate(
    Upgrade_Status = ifelse(is.na(UPGRADED), "Not Upgraded", "Upgraded")
  ) %>%
  group_by(Upgrade_Status) %>%
  summarise(count = n())  # Count the occurrences of each category

# Calculate percentages
moderate_comfort_data <- moderate_comfort_data %>%
  mutate(percentage = round((count / sum(count)) * 100, 2))  # Add percentage column

# Define colors for 'Upgraded' and 'Not Upgraded' categories
upgrade_colors_moderate <- c(
  "Upgraded" = "#8A7967",  # Gray-brown for upgraded
  "Not Upgraded" = "#B39B8E"  # Muted pinkish tone for not upgraded
)

# Generate the pie chart for Moderate Comfort lanes
ggplot(moderate_comfort_data, aes(x = "", y = percentage, fill = Upgrade_Status)) +
  geom_bar(stat = "identity", width = 1, color = "black") +  # Add black border lines around pie sections
  coord_polar("y", start = 0) +
  geom_text(aes(label = paste0(percentage, "%")), 
            position = position_stack(vjust = 0.5), 
            size = 6,   # Increase text size for percentage labels
            fontface = "bold") +  # Make percentage text bold
  scale_fill_manual(values = upgrade_colors_moderate) +  # Apply colors for Upgraded vs Not Upgraded
  labs(fill = "Upgrade Status") +
  theme_void() +  # Remove axis, grid lines, and background
  theme(legend.position = "right", 
        legend.title = element_text(size = 16),  # Increase legend title size
        legend.text = element_text(size = 14))  # Increase legend text size

# Filter data for 'High Comfort' lanes and categorize as 'Upgraded' or 'Not Upgraded'
high_comfort_data <- lane_type_data %>%
  filter(Comfort_Level == "High Comfort") %>%
  mutate(
    Upgrade_Status = ifelse(is.na(UPGRADED), "Not Upgraded", "Upgraded")
  ) %>%
  group_by(Upgrade_Status) %>%
  summarise(count = n())  # Count the occurrences of each category

# Calculate percentages
high_comfort_data <- high_comfort_data %>%
  mutate(percentage = round((count / sum(count)) * 100, 2))  # Add percentage column

# Define colors for 'Upgraded' and 'Not Upgraded' categories
upgrade_colors_high <- c(
  "Upgraded" = "#8A7967",  # Gray-brown for upgraded
  "Not Upgraded" = "#B39B8E"  # Muted pinkish tone for not upgraded
)

# Generate the pie chart for High Comfort lanes
ggplot(high_comfort_data, aes(x = "", y = percentage, fill = Upgrade_Status)) +
  geom_bar(stat = "identity", width = 1, color = "black") +  # Add black border lines around pie sections
  coord_polar("y", start = 0) +
  geom_text(aes(label = paste0(percentage, "%")), 
            position = position_stack(vjust = 0.5), 
            size = 6,   # Increase text size for percentage labels
            fontface = "bold") +  # Make percentage text bold
  scale_fill_manual(values = upgrade_colors_high) +  # Apply colors for Upgraded vs Not Upgraded
  labs(fill = "Upgrade Status") +
  theme_void() +  # Remove axis, grid lines, and background
  theme(legend.position = "right", 
        legend.title = element_text(size = 16),  # Increase legend title size
        legend.text = element_text(size = 14))  # Increase legend text size

@@

```{r}
#| label: fig-map
#| fig-cap: "Map of Bike Lanes Locations in Toronto"
#| echo: false
#| warning: false
#| message: false
#| fig-width: 11
#| fig-height: 8

# Load the data
coordinates_data <- read_csv(here("data", "analysis_data", "coordinates_data.csv"))

# Get bounding box for Toronto (approximate limits for Toronto)
toronto_bbox <- c(-79.6303856025883, 43.5822069220627, -79.1180336515019, 43.8554571861712)

# Fetch map data from OpenStreetMap (OSM)
toronto_map <- opq(bbox = toronto_bbox) %>%
  add_osm_feature(key = 'highway', value = 'primary') %>%
  osmdata_sf()

# Convert map data to Simple Features (sf) objects
streets <- toronto_map$osm_lines

# Plot the map with bike lanes data
ggplot() +
  geom_sf(data = streets, color = "gray", size = 0.3, alpha = 0.6) +  # OSM street map layer
  geom_point(data = coordinates_data, aes(x = longitude, y = latitude), 
             size = 0.1, color = "black") +  # Bike lanes points
  labs(x = "Longitude", y = "Latitude") +
  coord_sf(xlim = c(-79.6303856025883, -79.1180336515019), 
           ylim = c(43.5822069220627, 43.8554571861712), expand = FALSE) +
  theme_minimal() +
  theme(
    panel.background = element_blank(),  # Remove panel background
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

@fig-map provides a visual representation of the bike lane infrastructure across Toronto. The lines on the map correspond to various cycling paths, including dedicated bike lanes, multi-use paths, and shared roadways, which span different areas of the city. The geographic coordinates, ranging from around 43.60°N to 43.85°N in latitude and -79.6°W to -79.3°W in longitude, encompass most of Toronto’s urban region. This map offers an important spatial perspective, illustrating the extensive network of cycling routes that connect different neighborhoods and areas within the city.

A key observation from the map is the dense concentration of bike lanes in the downtown core and along major roadways, indicating a well-established network in central Toronto. The downtown area’s grid-like structure is particularly evident, showcasing a high density of lanes that facilitate easy cycling access within the city's busiest areas. As one moves further away from the city center, the map shows fewer bike lanes, suggesting that suburban regions may have less developed cycling infrastructure. However, there are notable multi-use paths extending toward the outskirts, particularly in parks and recreational areas, providing important cycling connections for those commuting or using bikes for recreational purposes.

This map highlights the importance of cycling infrastructure in promoting sustainable urban transportation. The wide distribution of bike lanes supports Toronto's efforts to enhance active transportation options, reduce traffic congestion, and improve accessibility for cyclists. The representation also provides a foundation for analyzing areas where future infrastructure developments could be prioritized, particularly in regions where cycling connectivity appears sparse.
